Probabilistic Catalogs for Crowded Stellar Fields

### 2013-03-03 paper/response.md

Cover Letter
============

Dear Dr. Gallagher,

Thank you for providing the referee's report on our manuscript AJ-11369
entitled "Probabilistic Catalogs for Crowded Stellar Fields". The
referee provided thoughtful commentary and constructive criticism of our
work, which was very valuable.

We have modified the manuscript significantly to address the concerns
of the referee. We list the changes below. We look forward to hearing the
decision on the revised manuscript.

@@

Brendon J. Brewer
Daniel Foreman-Mackey
David W. Hogg

Itemized Responses
==================

"Abstract seems to be overselling the case in multiple instances; for example, "introduce" may be too strong given Stetson's work on DAOPHOT"

- We have changed "introduce" to "present". Related approaches definitely exist
and we have added a few more citations that we have found about since (e.g.
Hobson 2003)

@@

"SExtractor "significantly underestimates" and "incorrect inferences"? What do "significant" and "incorrect" mean here - something of practical importance that astronomers should be scared of? The results presented in the paper didn't convince me."

- We were referring to the fact that SExtractor could not correctly infer the number of stars in the crowded case, even if you make the threshold settings very aggressive. This would be of practical importance in some instances but not others. It depends on whether finding all the stars or quantifying the faint stars is important, and that depends on what the scientist's goals are. The abstract has been clarified to be more specific and realistic about the differences between the methods.

@@

"The analysis of test cases is very superficial and the reader is left wondering about the actual performance. For example, what is the difference in photometric ans astrometric precision between the new method and SExtractor on an *object-by-object* basis? Surely at the bright end objects must be identical! What is completeness and contamination as a function of brightness? I return to this point below - something *must be wrong* with Fig. 6 and corresponding discussion. The bottom line is that currently Sections 5 and 6 are insufficient to assess the performance of this method, both in absolute sense and relative to SExtractor's performance. The unbelievable claims about SExtractor are not substantiated."

- We do not believe that there is anything wrong (technically) with Figure 6 or the discussion surrounding it (after extensive experimentation with parameters we have not been able to SExtract anywhere near 699 stars in Test Case 2). However, we realise that this choice of Figure is very dramatic and not particularly informative. Hence we have decided to replace Figure 6 with a different plot showing the true (cumulative) luminosity function along with inferences from our model and from SExtractor. We believe this plot shows more clearly what is going on, that the bright end is basically consistent (as one would expect) but that our method performs better at the faint end (although this is partially caused
by using a known parametric form for the LF, a point we now discuss in the paper). We have done some investigation of inferences about particular stars but the resulting plots were not very informative because sometimes, by chance, one or the other method did well or poorly on the selected star. The vertical dips in the CLF plot show similar information in any case.

@@

"- p. 2: PHOTO should be photo"

- Fixed

"p. 2, first paragraph: Can't you simplify this long discussion by just stating that catalogs are lists of best-fit model parameters?"

- Sometimes they are lists of best fitting model parameters (by some criterion of "best"), but sometimes they are not. It depends on how they were produced. SExtractor, for example, doesn't really resembling fitting. It's a point estimation procedure.

@@

"very lossy", "destroy information" etc. - this makes it sound like the end of the world. Either be quantitative or add appropriate caveats; (overselling this method may be counterproductive).

- In principle it is of course lossy and destroys information, but the degree to which this is practically important is not clear yet. We have moderated our language to this effect.

@@

"p. 2, third paragraph: I am *very* confused by the statement that "Instead of estimating..." First, there is nothing more than "estimating" that we can do whatever the method. Second, and more importantly, standard methods (as in SExtractor, photo, and most other codes) *do* estimate errors for flux, positions, and many other parameters, and these error estimates *are* propagated (in most papers) to subsequent analysis. What is *not* probabilistic about giving flux+-error?"

- This is a language issue. We were using estimation in the statistics sense of "point estimates obtained from some procedure". Standard errors are related to probability but there are many ways of obtaining them in the community.

@@

"general comment on Introduction: typically, one knows what exactly the paper is about after reading it - not the case here."

- The introduction has been changed somewhat, hopefully making this more clear.

@@

"section 3.3: it is a major omission not to give credit to Stetson and DAOPHOT here"

- After rereading Stetson it is clear that the approach is related. We have added a citation here.

@@

"fig. 4: pretty but useless! What is the reader supposed to do - stare at it and count all the points? It would be better to use images such as Fig. 1 and 2, and show the mean image for all catalogs, and then a few residual images (perhaps normalized by statistical errors)."

- We do show these summary images in Figure 5. We would like to retain Figure 4 primarily because it shows the fundamental output of this approach, which is samples, not point estimates. We also like the fact that it is pretty!

"fig. 5: what are the stretch and scale in the lower left panel? How do residuals compare to statistical fluctuations?"

- The residuals are just the statistical fluctuations and pass all traditional tests such as reduced chi-squared. This essentially has to be true because we know the model is correct. In practice this may not be the case on real data, in which
case the model assumptions (PSF etc) would need to be modified.

@@

"Section 6: Fig. 6 is very misleading (it implies that SExtractor, which was so many times successfully tested, is lousy). A much more informative plot (I tried it in practice in a similar context) would be to plot cumulative counts vs. log(flux) (bright to faint) for your method and for SExtractor on the same panel. Additional useful plot would be to match sources from the two lists within some small radius and compare their photometry and astrometry with true values. It wouldn't be surprising if SExtractor has biases in crowded field case, and such biases would be *a really strong* argument in favor of your method!"

- We have replaced Figure 6 with the suggested luminosity function plot. It shows that the SExtractor catalogs lie within the uncertainties from the posterior distribution but *only* at the bright end, as you suspect. Our method outperforms SExtractor at the faint end. This can be partially explained by our use of a known parametric form for the LF. To clarify this we have added some discussion of this point.

"Section 6: running SExtractor down to 0.5sigma and computing LF with all these sources is pretty silly (as the sample will have a lot of faint junk; also large errors and Malmquist bias will be important). I find the clustering of alpha1 ~ alpha2 ~ 1 strongly suspect. Also, even for a clean sample with vanishing errors, there are many pitfalls when fitting power-law LFs. I strongly suggest the authors take a look at Clauset, Shalizi, and Newman (2009). Power-Law Distributions in Empirical Data. SIAM Review 51, 661-703."

- I do not understand how else one would compute the LF from a traditional catalog apart from computing the catalog and then fitting the power-law. We are aware of the pitfalls of power laws and have added a caveat saying that it's mostly for demonstration purposes. However, we have removed this figure because it is too dramatic.

@@

"Discussion: second paragraph on p. 15: I don't think your model is limited for what you wanted to do - that is, to assess your method from a statistical point of view. What you listed are limitations for applying your procedure to real data."

- Correct. We have modified the conclusions with the aim of separating out future testing-work (comparing to more codes like DAOPHOT and DOLPHOT) from realistic-science work (which we discussed here).

### 2013-03-03 paper/ms.tex
Paper tweaks

\begin{abstract}
We present and implement a probabilistic (Bayesian) method for producing catalogs
from images of stellar fields. The method is capable of inferring the
number of sources $N$ in the image and can also handle the
challenges introduced by noise and overlapping sources. The luminosity
function of the stars can also be inferred even when the precise luminosity
of each star is uncertain, via the use of a hierarchical Bayesian model.
The computational feasibility of the method is
demonstrated on two simulated images with different numbers of stars.
We find that our method
successfully recovers the input parameter values along with principled
uncertainties even when the field is crowded.
We also compare our results with those obtained from the
\sex~software. While the two approaches largely agree about the fluxes of the
bright stars, the Bayesian approach provides more accurate inferences about
the faint stars and the number of stars, particularly in the crowded case.
\end{abstract}

@@

\section{Introduction}
Traditional practice in astronomy is to take images of the sky, detect
or enumerate sources visible in those images, and create catalogs.
These catalogs are then used to perform the fundamental astronomical
measurements, which might include something like the three-dimensional
structure of the Galaxy or the two-point correlation function of
galaxies.  Indeed, the process of catalog construction is so ``baked
in'' to our ideas about what astronomy is, we sometimes forget that
the catalog is \emph{not} the fundamental data product of astronomy; catalogs are produced
from imaging; their production involves many decisions and ideas
that go way beyond the information provided to the telescope by the
incident intensity field. In addition, catalogs are not usually the final
goal of any imaging project or survey. Typically, they are produced in
order to facilitate the scientific study of populations of objects (e.g. the
initial mass function of a population of stars), or to provide a sky-search
capability to the community who might be interested in only a small subset of
objects. Standard tools for generating catalogs from astronomical imaging
include \sex~\citep{sextractor}, DAOPHOT \citep{1987PASP...99..191S}, 
DOLPHOT \citep{dolphot}, and SDSS
Photo \citep{photo}.

Telescopes don't make catalogs \citep{2011EAS....45..351H}, they
measure the intensity field.  Viewed through the lens of probabilistic
inference, the goals of astronomy are to take the information in the
telescope-generated records of the intensity field and deliver it to
the quantities of astronomical interest with as little loss as
possible.  Insertion of a catalog-generation step in the inference
pipeline between the raw imaging and the final astrophysical analyses
is potentially lossy; the hard decisions of catalog making destroy
information, at least in principle.  Probability theory suggests that it may be less
lossy to pass forward not a catalog but a probabilisitic
description of all the catalogs that could be consistent with the
imaging---a posterior probability in (enormously large) space of possible catalogs.
Essentially, the creation of a catalog is an
attempt to answer the question,
``Given the image we have obtained, what objects are present in the field and
what are their properties?''.
This article represents an attempt at this ambitous goal in the specific
situation where the only objects in the field are stars or other point sources.

Beyond these philosophical concerns, there are practical issues:
Standard methods for constructing catalogs can have difficulty in some
challenging situations. For example, when multiple sources overlap partially
or completely, it can be difficult to determine how many sources are present,
and how much flux belongs to each source. In principle, the uncertainty about
the existence and properties of the stars can be significant and should be
propagated into any inferences about the stellar population. A Bayesian approach
that obtains the posterior distribution over catalog space (rather than a single
catalog estimate) has the potential to overcome these problems by deblending
objects when it is possible, and clearly indicating the uncertainty remaining
when it is not possible.

Sampling a posterior probability distribution for catalogs is a challenging
numerical task for a number of reasons. Firstly, the number $N$ of objects in the
image (and that should therefore be listed in the catalog) is itself unknown.
Secondly, if $N$ is large, then the parameter space
of positions and properties (flux, size, etc) of the objects is also large.
This can cause Markov Chain Monte Carlo (MCMC) algorithms difficulties -- they
may take a long time to converge to the target posterior distribution over
the space of catalogs. Thirdly, this problem is subject to the so-called
label-switching problem that is commonly encountered in mixture modeling
\citep[e.g.][]{label_switching}. Given any proposed catalog, another catalog that is
equally plausible is the catalog obtained by shuffling the entries of the first
catalog. This leads to a posterior distribution with $N!$ identical peaks in
parameter space. This can lead to difficulties with certain (generally very
effective) MCMC algorithms such as the affine-invariant stretch move
\citep{goodman, emcee}.

@@

In this paper, we develop a Bayesian object detection model with the following
features: i) the number $N$ of objects in the image is an unknown parameter
to be inferred, ii) the objects that we expect to find are point sources such
as stars, and iii) the point-spread function is unknown (but a parametric model is
used) and must be inferred from
the data (but a single bright star may not be available to help with estimating it).
This paper is structured as follows. In Section~\ref{sec:bayes}
we give a brief overview of Bayesian inference. In Section~\ref{sec:model} we
discuss the model assumptions we make in our method. In Section~\ref{sec:mcmc}
we briefly discuss our
MCMC implementation. Section~\ref{sec:simulated_data} describes the tests we
carried out on simulated data, and a comparison with \sex~results is presented
in Section~\ref{sec:sex}. We conclude in Section~\ref{sec:conclusion}.

@@

Throughout this paper we will
assume the pixel-convolved PSF is a weighted mixture of two concentric
circular
Gaussians (a similar mixture model is also used by DAOPHOT, \citet{1987PASP...99..191S}) with widths
$s_1$ and $s_2$:
\begin{eqnarray}
\mathcal{P}(x, y) = \frac{w}{2\pi s_1^2}\exp
\left[
-\frac{1}{2s_1^2}\left(x^2 + y^2\right)
\right]
+ \frac{1-w}{2\pi s_2^2}\exp
\left[-\frac{1}{2s_2^2}\left(x^2 + y^2\right)
\right].
\end{eqnarray}
The pixellated observed image is assumed to be generated from the mock
image (evaluated at the center of each pixel) plus noise:
\begin{eqnarray}
I_{ij} &=& \mathcal{M}(X_{ij}, Y_{ij}) + \epsilon_{ij}
\end{eqnarray}
where the errors $\{\epsilon_{ij}\}$ are independent and normally distributed.
The variance of the normal distribution for each pixel is determined by the
brightness of the sky and the
brightness of the mock image in that pixel. This can be modelled by assuming the
following distribution:
\begin{eqnarray}
\epsilon_{ij} \sim \mathcal{N}(0, \sigma_0^2 + \eta \mathcal{M}(X_{ij}, Y_{ij})).
\end{eqnarray}
where $\sigma_0$ is a constant noise level and $\eta$ is a coefficient allowing
the noise to be higher in brighter regions of the image. This dependence of the
noise variances on the model intensity arises as a result of the Poissonian
nature of photon counts, but allows for the fact that a ``sky'' background may
have already been subtracted from the image in the reduction process.
This parameterisation has been used by \citet{2011MNRAS.412.2521B} and is an
alternative to the common practice of producing a ``variance map'' from the
image data that is then assumed to be known.

@@

\section{Comparison to SExtractor}\label{sec:sex}
In the previous section we established that the inference of the catalogs from
the data is computationally feasible and that the number of stars and the
luminosity function can be inferred from the image data, albeit with moderate
uncertainties. We now compare this approach to an alternative analysis that makes
use of the standard
tool \sex~\citep{sextractor}. To achieve this, we executed \sex~on the two test images, for various
values of the detection threshold parameters {\tt DETECT\_THRESH} and {\tt ANALYSIS\_THRESH}
ranging from 0.5 to 6.5. This results in a set of catalogs
for each image, with more conservative thresholds resulting in less stars detected
as compared to more aggressive thresholds. The results for the number of stars in
the images are shown in Figure~\ref{fig:luminosity_function}. To compute flux
estimates that are directly comparable to the fluxes in our input catalogs, we
configured \sex~to compute object fluxes within fixed circular apertures that were
known to contain 70\% of the mass of the PSF. The \sex~flux estimates were then
scaled up to account for this finite aperture.

>>>>

\begin{abstract}
We present and implement a probabilistic (Bayesian) method for producing catalogs
from images of stellar fields. The method is capable of inferring the
number of sources $N$ in the image and can also handle the
challenges introduced by noise, overlapping sources, and an unknown
point spread function (PSF). The luminosity
function of the stars can also be inferred even when the precise luminosity
of each star is uncertain, via the use of a hierarchical Bayesian model.
The computational feasibility of the method is
demonstrated on two simulated images with different numbers of stars.
We find that our method
successfully recovers the input parameter values along with principled
uncertainties even when the field is crowded.
We also compare our results with those obtained from the
\sex~software. While the two approaches largely agree about the fluxes of the
bright stars, the Bayesian approach provides more accurate inferences about
the faint stars and the number of stars, particularly in the crowded case.
\end{abstract}

@@

\section{Introduction}
Traditional practice in astronomy is to take images of the sky, detect
or enumerate sources visible in those images, and create catalogs.
These catalogs are then used to perform the fundamental astronomical
measurements, which might include something like the three-dimensional
structure of the Galaxy or the two-point correlation function of
galaxies.  Indeed, the process of catalog construction is so ``baked
in'' to our ideas about what astronomy is, we sometimes forget that
the catalog is \emph{not} the fundamental data product of astronomy; catalogs are produced
from imaging; their production involves many decisions and ideas
that go beyond the information provided to the telescope by the
incident intensity field. In addition, catalogs are not usually the final
goal of any imaging project or survey. Typically, they are produced in
order to facilitate the scientific study of populations of objects (e.g. the
initial mass function of a population of stars), or to provide a sky-search
capability to the community who might be interested in only a small subset of
objects. Standard tools for generating catalogs from astronomical imaging
include \sex~\citep{sextractor}, DAOPHOT \citep{1987PASP...99..191S}, 
DOLPHOT \citep{dolphot}, and SDSS
Photo \citep{photo}.

Telescopes don't make catalogs \citep{2011EAS....45..351H}, they
measure the intensity field.  Viewed through the lens of probabilistic
inference, the goals of astronomy are to take the information in the
telescope-generated records of the intensity field and use this information
to obtain quantities of astronomical interest with as little loss as
possible.  Insertion of a catalog-generation step in the inference
pipeline between the raw imaging and the final astrophysical analyses
is potentially lossy; the hard decisions of catalog making destroy
information, at least in principle.  Probability theory suggests that it may be less
lossy to pass forward not a catalog but a probabilisitic
description of all the catalogs that could be consistent with the
imaging---a posterior probability in (enormously large) space of possible catalogs.
Essentially, the creation of a catalog is an
attempt to answer the question,
``Given the image we have obtained, what objects are present in the field and
what are their properties?''.
This article represents an attempt at implementing this ambitous goal in the specific
situation where the only objects in the field are stars or other point sources.

Beyond these philosophical concerns, there are practical issues:
Standard methods for constructing catalogs can have difficulty in some
challenging situations. For example, when multiple sources overlap partially
or completely, it can be difficult to determine how many sources are present,
and how much flux belongs to each source. In principle, the uncertainty about
the existence and properties of the objects can be significant and should be
propagated into any inferences about the stellar population. A Bayesian approach
that obtains the posterior distribution over catalog space (rather than a single
catalog estimate) has the potential to overcome these problems by deblending
objects when it is possible, and clearly indicating the uncertainty remaining
when it is not possible.

In practice, Bayesian Inferences are often implemented using Markov Chain Monte
Carlo (MCMC) methods \citep{mackay} to sample from the posterior distribution.
Sampling a posterior probability distribution for catalogs is a challenging
numerical task for a number of reasons. Firstly, the number $N$ of objects in the
image (and that should therefore be listed in the catalog) is itself unknown.
Secondly, if $N$ is large, then the parameter space
of positions and properties (flux, size, etc) of the objects is also large.
This can cause Markov Chain Monte Carlo (MCMC) algorithms difficulties -- they
may take a long time to converge to the target posterior distribution over
the space of catalogs. Thirdly, this problem is subject to the so-called
label-switching problem that is commonly encountered in mixture modeling
\citep[e.g.][]{label_switching}. Given any proposed catalog, another catalog that is
equally plausible is the catalog obtained by shuffling the entries of the first
catalog. This leads to a posterior distribution with $N!$ identical peaks in
parameter space. This can lead to difficulties with certain (otherwise very
effective) MCMC algorithms such as the affine-invariant stretch move
\citep{goodman, emcee}.

@@

In this paper, we develop a Bayesian object detection model with the following
features: i) the number $N$ of objects in the image is an unknown parameter
to be inferred from the data, ii) the objects that we expect to find are point sources such
as stars, and iii) the point-spread function is unknown (but a parametric model is
used) and must be inferred from
the data (but a single bright star may not be available to help with estimating it).
This paper is structured as follows. In Section~\ref{sec:bayes}
we give a brief overview of Bayesian inference. In Section~\ref{sec:model} we
discuss the model assumptions we make in our method. In Section~\ref{sec:mcmc}
we briefly discuss our
MCMC implementation. Section~\ref{sec:simulated_data} describes the tests we
carried out on simulated data, and a comparison with \sex~results is presented
in Section~\ref{sec:sex}. We conclude in Section~\ref{sec:conclusion}.

@@

Throughout this paper we will
assume the pixel-convolved PSF is a weighted mixture of two concentric
circular
Gaussians (a similar mixture model that uses a Gaussian core and flexible wings is also used by DAOPHOT, \citet{1987PASP...99..191S}) with widths
$s_1$ and $s_2$:
\begin{eqnarray}
\mathcal{P}(x, y) = \frac{w}{2\pi s_1^2}\exp
\left[
-\frac{1}{2s_1^2}\left(x^2 + y^2\right)
\right]
+ \frac{1-w}{2\pi s_2^2}\exp
\left[-\frac{1}{2s_2^2}\left(x^2 + y^2\right)
\right].
\end{eqnarray}
The pixellated observed image is assumed to be generated from the mock
image (evaluated at the center of each pixel) plus noise:
\begin{eqnarray}
I_{ij} &=& \mathcal{M}(X_{ij}, Y_{ij}) + \epsilon_{ij}
\end{eqnarray}
where the errors $\{\epsilon_{ij}\}$ are independent and normally distributed.
The variance of the normal distribution for each pixel is determined by the
brightness of the sky and the
brightness of the mock image in that pixel. This can be modelled by assuming the
following distribution:
\begin{eqnarray}
\epsilon_{ij} \sim \mathcal{N}(0, \sigma_0^2 + \eta \mathcal{M}(X_{ij}, Y_{ij})).
\end{eqnarray}
where $\sigma_0$ is a constant noise level and $\eta$ is a coefficient allowing
the noise to be higher in brighter regions of the image. This dependence of the
noise variances on the model intensity arises as a result of the Poissonian
nature of photon counts, but allows for the fact that a ``sky'' background may
have already been subtracted from the image in the reduction process.
This parameterisation has been used by \citet{2011MNRAS.412.2521B} and is an
alternative to the common practice of producing a ``variance map'' from the
image data that is then assumed to be known.

@@

\section{Comparison to SExtractor}\label{sec:sex}
In the previous section we established that the inference of the catalogs from
the data is computationally feasible and that the number of stars and the
luminosity function can be inferred from the image data, albeit with moderate
uncertainties. We now compare this approach to an alternative analysis that makes
use of the standard
tool \sex~\citep{sextractor}. To achieve this, we executed \sex~on the two test images, for various
values of the detection threshold parameters {\tt DETECT\_THRESH} and {\tt ANALYSIS\_THRESH}
ranging from 0.5 to 6.5. This results in a set of catalogs
for each image, with more conservative thresholds resulting in less stars detected
as compared to more aggressive thresholds. To compute flux
estimates that are directly comparable to the fluxes in our input catalogs, we
configured \sex~to compute object fluxes within fixed circular apertures that were
known to contain 70\% of the mass of the PSF. The \sex~flux estimates were then
scaled up to account for this finite aperture.
